---
title: "Derieved columns in pepr"
author: "Michal Stolarczyk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will show you how and why to use the derieved columns functionality of the `pepr` package. For introductory information about the package visit the [documentation page](http://code.databio.org/pepr/index.html). For the basic information about the PEP concept visit the [project website](https://pepkit.github.io/). For a broader theoretical description of derived columns click [here](https://pepkit.github.io/docs/derived_columns/).

### Problem/Goal
The example below demonstrates how to use the derived columns to **flexibly define the samples attributes the `file_path` column** of the `sample_annotation.csv` file to match the file names in  your project. Please consider the example below for reference:


```{r ,echo=FALSE}
library(knitr)
library(kableExtra)
sampleAnnotation = system.file("extdata", "example_derived", "sample_annotation_pre.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
knitr::kable(sampleAnnotationDF,format = "html") 
```

----
### Solution
As the name suggests the attributes in the specified columns (here: `file_path`) can be derived from other ones. The way how this process is carried out is indicated explicitly in the `project_config.yaml` file (presented below). The name of the column is determined in the `derived_columns` key-value pair, whereas the pattern for the attributes construction - in the `data_sources` one. Note that the second level key (here: `source`) has to exactly match the attributes in the `file_path` column of the modified `sample_annotation.csv` (presented below).

```{yaml, highlight=TRUE}
metadata:
    sample_annotation: sample_annotation.csv
    output_dir: $HOME/hello_looper_results

derived_columns: [file_path]
data_sources:
    source1: /data/lab/project/{organism}_{time}h.fastq
    source2: /path/from/collaborator/weirdNamingScheme_{external_id}.fastq
```

Let's introduce a few modifications to the original `sample_annotation.csv` file to map the appropriate data sources from the `project_config.yaml` with attributes in the derived column - `[file_path]`:

```{r ,echo=FALSE}
library(knitr)
library(kableExtra)
sampleAnnotation = system.file("extdata", "example_derived", "sample_annotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
----
### Code
Load `pepr` and read in the project metadata by specifying the path to the `project_config.yaml`:

```{r}
library(pepr)
projectConfig = system.file("extdata", "example_derived", "project_config.yaml", package="pepr")
p=Project(projectConfig)
```

And inspect it:

```{r}
p@samples
```

As you can see, the resulting samples are annotated the same way as if they were read from the original, unwieldy, annotations file.

What is more, the `p` object consists of all the information from the project config file (`project_config.yaml`). Run the following line to explore it:

```{r}
p@config
```


