---
title: "pepr - SummarizedExperiment interface"
author: "Michal Stolarczyk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction

The [`pepr`](https://pepkit.github.io/docs/pepr/) package implements a class that provides an interface for the `Project` and [`SummarizedExperiment`](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) classes. This way the users of `SummarizedExperiment` can benefit from the utilties and ideology of the [PEP](https://pepkit.github.io/) toolkit. Such as: 

* description of all projects the same way, so you only have to learn one metadata structure
* collaborative work, since the organization is standardized
* sharing the project with others, who are already familiar with the standard
* employment of a variety of tools or pipelines without restructuring the metadata
* analyis attributes of your project in the data analysis environment of your choice (R/python).

### How to create the interface?

Load the package
```{r echo=T,message=FALSE,warning=FALSE}
# Loading of packages and sourcing the file with the interface class will be substituded by loading of the final package in the future
library(pepr)
library(SummarizedExperiment)
source(file = "../interface_class.R")

```
Let's create an empty `Interface` and see how it looks like
```{r}
i = Interface()
i
slotNames(i)
```
As you can see, `inter` the object of class `Interface` consists of the attributes of the `SummarizedExperiment` objects (such as `assays`, `colData` etc.) and an object of class `Project` that is an in-memory representation of a PEP.

Obviously, an empty object of `Interface` class is useless, so let's consider the following example:

1. Generate the data for `SummarizedExperiment`

```{r}
nrows = 200
ncols = 6
counts = matrix(runif(nrows * ncols, 1, 1e4), nrows)
rowRanges = GRanges(rep(c("chr1", "chr2"), c(50, 150)),
                    IRanges(floor(runif(200, 1e5, 1e6)), width=100),
                    strand=sample(c("+", "-"), 200, TRUE),
                    feature_id=sprintf("ID%03d", 1:200))
colData = DataFrame(Treatment=rep(c("ChIP", "Input"), 3),
                    row.names=LETTERS[1:6])

```

2. Generate an object of class `Project` that holds a PEP.

```{r}
projectConfig = system.file(
"extdata",
"example_peps-master",
"example_implied",
"project_config.yaml",
package = "pepr"
)
pep=Project(projectConfig)
```

3. Create the `Interface` object and inspect it

```{r}
i = Interface(pep=pep, assays=list(counts=counts), rowRanges=rowRanges, colData=colData)
i

```

### How to integrate the `SummarizedExperiment` and `Project`?

Currently, there are 2 methods that can integrate the data from `Project` object with the `SummarizedExperiment` within the `Interface` obejcts: `getColData()` and `getMetadata()`.

The former one copies the data about samples from `samples` slot in objects of `Project` class to the `colData` slot in the `SummarizedExperiment` objects. See the example below:

The current (generated) `colData`:

```{r}
colData(i)
```

Copying the colData with the `Interface` methods:

```{r}
iModified = getColData(i)
colData(iModified)
```

Please note, that the "getters and setters" methods associated with the `SummarizedExperiment` class (`rowRanges()`, `rowData()`, `colData()`, `metadata()`) are inherited and work with the `Interface` class objects.

The `getMetadata()` method enriches the current `metadata` section of `SummarizedExperiment` with the path to the config file that was used to create the `Project` obejct (PEP).

The current (generated) `metadata`:

```{r}
metadata(i)
```

Enriched `metadata`:

```{r}
iModified = getMetadata(i)
metadata(iModified)
```


