---
title: "Sample subannotations in pepr"
author: "Michal Stolarczyk & Nathan Sheffield"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will show you how and why to use the sample subannotation functionality of the `pepr` package. For introductory information about the package visit the [documentation page](http://code.databio.org/pepr/index.html). For the basic information about the PEP concept visit the [project website](https://pepkit.github.io/). For a broader theoretical description of sample subannotation click [here](https://pepkit.github.io/docs/sample_subannotation/).

### Problem/Goal

This series of examples below demonstrates how and why to use sample subannoatation functionality in multiple cases to **provide multiple input files of the same type for a single sample**.

### Solutions

#### Example 1: basic sample subannotation table

This example demonstrates how the sample subannotation functionality is used. In this example, 2 samples have multiple input files that need merging (`frog_1` and `frog_2`), while 1 sample (`frog_3`) does not. Therefore, `frog_3` specifies its file in the `sample_annotation.csv` file, while the others leave that field blank and instead specify several files in the `sample_subannotation.csv` file.

This example is made up of these components:

* Project config file:
```{r, echo=FALSE,message=TRUE,collapse=TRUE,comment=" "}
library(pepr)
project_config = system.file("extdata","example_peps-master", "example_subannotation1", "project_config.yaml", package="pepr")
printNestedList(yaml::read_yaml(project_config))
```
* Sample annotation table:
```{r ,echo=FALSE}
library(knitr)
sampleAnnotation = system.file("extdata","example_peps-master", "example_subannotation1", "sample_annotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
* Sample subannotation table:
```{r ,echo=FALSE}
sampleAnnotation = system.file("extdata","example_peps-master", "example_subannotation1", "sample_subannotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```

Let's load the project config, create the Project object and see if multiple files are present 
```{r}
library(pepr)
projectConfig1=system.file("extdata", "example_peps-master","example_subannotation1", "project_config.yaml", package="pepr")
p1=Project(projectConfig1)
# Check the files
p1@samples$file
# Check the subsample names
p1@samples$subsample_name
```
And inspect the whole table in `p1@samples` slot
```{r,echo=FALSE}
kable(p1@samples)
```

You can also access a single subsample if you call the `getSubsample` method with appropriate `sample_name` - `subsample_name` attribute combination. Note, that this is only possible if the `subsample_name` column is defined in the `sub_annotation.csv` file.

```{r}
sampleName="frog_1"
subsampleName="sub_a"
getSubsample(p1,sampleName,subsampleName)
```


#### Example 2: subannotations and derived columns

This example uses a `sample_subannotation.csv` file and a derived column to point to files. This is a rather complex example. Notice we must include the `file_id` column in the `sample_annotation.csv` file, and leave it blank; this is then populated by just some of the samples (`frog_1` and `frog_2`) in the `sample_subannotation.csv`, but is left empty for the samples that are not merged.

This example is made up of these components:

* Project config file:
```{r, echo=FALSE,message=TRUE,collapse=TRUE,comment=" "}
library(pepr)
project_config = system.file("extdata", "example_peps-master","example_subannotation2", "project_config.yaml", package="pepr")
printNestedList(yaml::read_yaml(project_config))
```
* Sample annotation table:
```{r ,echo=FALSE}
library(knitr)
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation2", "sample_annotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
* Sample subannotation table:
```{r ,echo=FALSE}
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation2", "sample_subannotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
Let's load the project config, create the Project object and see if multiple files are present 
```{r}
library(pepr)
projectConfig2=system.file("extdata", "example_peps-master","example_subannotation2", "project_config.yaml", package="pepr")
p2=Project(projectConfig2)
# Check the files
p2@samples$file
```
And inspect the whole table in `p2@samples` slot
```{r,echo=FALSE}
kable(p2@samples)
```

#### Example 3: subannotations and expansion characters

This example gives the exact same results as Example 2, but in this case, uses a wildcard for `frog_2` instead of including it in the `sample_subannotation.csv` file. Since we can't use a wildcard and a subannotation for the same sample, this necessitates specifying a second data source class (`local_files_unmerged`) that uses an asterisk (*). The outcome is the same.

This example is made up of these components:

* Project config file:
```{r, echo=FALSE,message=TRUE,collapse=TRUE,comment=" "}
library(pepr)
project_config = system.file("extdata", "example_peps-master","example_subannotation3", "project_config.yaml", package="pepr")
printNestedList(yaml::read_yaml(project_config))
```
* Sample annotation table:
```{r ,echo=FALSE}
library(knitr)
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation3", "sample_annotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
* Sample subannotation table:
```{r ,echo=FALSE}
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation3", "sample_subannotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
Let's load the project config, create the Project object and see if multiple files are present 
```{r}
library(pepr)
projectConfig3=system.file("extdata", "example_peps-master","example_subannotation3", "project_config.yaml", package="pepr")
p3=Project(projectConfig3)
# Check the files
p3@samples$file
```
And inspect the whole table in `p3@samples` slot
```{r,echo=FALSE}
kable(p3@samples)
```

#### Example 4: subannotations and multiple (separate-class) inputs

Merging is for same class inputs (like, multiple files for read1). Different-class inputs (like read1 vs read2) are handled by different attributes (or columns). This example shows you how to handle paired-end data, while also merging within each.

This example is made up of these components:

* Project config file:
```{r, echo=FALSE,message=TRUE,collapse=TRUE,comment=" "}
library(pepr)
project_config = system.file("extdata", "example_peps-master","example_subannotation4", "project_config.yaml", package="pepr")
printNestedList(yaml::read_yaml(project_config))
```
* Sample annotation table:
```{r ,echo=FALSE}
library(knitr)
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation4", "sample_annotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
* Sample subannotation table:
```{r ,echo=FALSE}
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation4", "sample_subannotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
Let's load the project config, create the Project object and see if multiple files are present 
```{r}
library(pepr)
projectConfig4=system.file("extdata", "example_peps-master","example_subannotation4", "project_config.yaml", package="pepr")
p4=Project(projectConfig4)
# Check the read1 and read2 columns
p4@samples$read1
p4@samples$read2
```
And inspect the whole table in `p4@samples` slot
```{r,echo=FALSE}
kable(p4@samples)
```

#### Example 5: subannotations and multiple (separate-class) inputs with derived columns

Merging is for same class inputs (like, multiple files for read1). Different-class inputs (like read1 vs read2) are handled by different attributes (or columns). This example shows you how to handle paired-end data, while also merging within each. Additionally, in this example the derived columns are used to create `read1` and `read2` attributes.

This example is made up of these components:

* Project config file:
```{r, echo=FALSE,message=TRUE,collapse=TRUE,comment=" "}
library(pepr)
project_config = system.file("extdata", "example_peps-master","example_subannotation5", "project_config.yaml", package="pepr")
printNestedList(yaml::read_yaml(project_config))
```
* Sample annotation table:
```{r ,echo=FALSE}
library(knitr)
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation5", "sample_annotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
* Sample subannotation table:
```{r ,echo=FALSE}
sampleAnnotation = system.file("extdata", "example_peps-master","example_subannotation5", "sample_subannotation.csv", package="pepr")
sampleAnnotationDF=read.table(sampleAnnotation,sep = ",",header = T)
kable(sampleAnnotationDF,format = "html") 
```
Let's load the project config, create the Project object and see if multiple files are present 
```{r}
library(pepr)
projectConfig5=system.file("extdata", "example_peps-master","example_subannotation5", "project_config.yaml", package="pepr")
p5=Project(projectConfig5)

# Check the read1 and read2 columns
p5@samples$read1
p5@samples$read2
```
And inspect the whole table in `p5@samples` slot
```{r,echo=FALSE}
kable(p5@samples)
```
